---
title: "Step-by-Step Statistical Analysis via SQL: Advanced Median & Anomalies"
author: "Veriwa Elisha"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
---

This part of the portfolio is a continuation and in this section, I attempt to tackle a very common real-world scenario: **numeric variables stored as text (strings) mixed with anomalies**. 

If you try to calculate a median on a column containing `"15.5 mins"`, `"N/A"`, or `"1.2.3"`, standard SQL will throw an error. I attempt using PostgreSQL's robust regex and casting functions to extract valid numbers, identify extreme outliers using the IQR method, and *then* calculate the median.

```{r setup-advanced, include=FALSE}
library(DBI)
library(duckdb)
library(ggplot2)
library(dplyr)

# Initialise in-memory DuckDB (PostgreSQL dialect)
con <- dbConnect(duckdb())

# Helper to run SQL and return a dataframe
run_sql <- function(query) {
  dbGetQuery(con, query)
}

theme_set(theme_minimal() + theme(plot.title = element_text(face = "bold")))
set.seed(842)
```

## 🎯 The Objective: Medians with Messy Strings & Outliers

In production databases, numeric data is often incorrectly logged as strings (e.g., `"45 mins"`, `"N/A"`, `"12.5.5"`). Furthermore, extreme outliers (like a bot typing 9999) can skew the mathematical thresholds used to identify anomalies. 

In this section, we will:
1. Extract valid numerics from a poorly formatted text column using SQL Regex.
2. Use the **Interquartile Range (IQR) method** in SQL to define what constitutes an "anomaly."
3. Filter those anomalies and calculate a mathematically sound **Median**.

---

## Step 7: Simulating Incorrect Formats & Extreme Anomalies

### Step 7.1: Generate a new messy text-based numeric column
```{r}
# Let's grab the clean_events table from the previous analysis, but destroy the formatting
# of session_time_clean to simulate a raw, badly logged text column.
messy_text_data <- run_sql("SELECT user_id, group_assignment, revenue FROM clean_events")

# Inject strings, multiple decimals, and extreme outliers
messy_text_data$session_text <- sapply(messy_text_data$session_time_clean <- runif(nrow(messy_text_data), 5, 25), function(x) {
  bug_type <- sample(1:5, 1, prob = c(0.7, 0.1, 0.1, 0.05, 0.05))
  if(bug_type == 1) return(paste0(round(x, 1), " mins"))
  if(bug_type == 2) return(paste0(round(x, 2), "minutes"))
  if(bug_type == 3) return("N/A")
  if(bug_type == 4) return(paste0(x, ".", sample(10:99, 1))) # Multiple decimals: 15.34.45
  if(bug_type == 5) return(as.character(sample(500:1000, 1))) # Extreme bot outlier
})

# Write to database
dbWriteTable(con, "raw_text_events", messy_text_data, overwrite = TRUE)
```

---

## Step 8: Diagnosing the Incorrect Formats

Before calculating anything, we must profile the text column to understand the data quality.

### Step 8.1: Inspect raw string values
```{r}
run_sql("
SELECT session_text, COUNT(*) AS count
FROM raw_text_events
GROUP BY session_text
ORDER BY count DESC
LIMIT 10;
")
```

---

## Step 9: Extracting Valid Numbers with SQL Regex

We need to isolate the first valid number from the string. PostgreSQL allows us to use `REGEXP_REPLACE` or `SUBSTRING`. Here, we will strip out everything except digits and decimals, then use `TRY_CAST` to convert it to a numeric format safely (invalid conversions become `NULL`).

### Step 9.1: Clean and cast the text to numeric
```{r}
run_sql("
CREATE OR REPLACE TABLE extracted_numeric AS
SELECT 
  user_id,
  group_assignment,
  revenue,
  -- Step 1: Keep only digits and dots. 
  -- Step 2: TRY_CAST to numeric. If it fails (e.g., multiple dots), it becomes NULL.
  TRY_CAST(REGEXP_REPLACE(session_text, '[^0-9.]', '', 'g') AS DOUBLE) AS session_numeric
FROM raw_text_events;

SELECT session_text, session_numeric
FROM raw_text_events r
JOIN extracted_numeric e USING(user_id)
WHERE r.session_text LIKE '%N/A%' 
   OR r.session_text LIKE '%mins%'
LIMIT 10;
")
```

---

## Step 10: Identifying Anomalies using the IQR Method

Now that we have valid numbers, we still have extreme outliers (e.g., 500+ minutes). Standard deviation won't work well here because the outliers inflate the mean and SD. Instead, we use the **Interquartile Range (IQR)**.

**Formula:** An anomaly is anything below `Q1 - (1.5 * IQR)` or above `Q3 + (1.5 * IQR)`.

### Step 10.1: Calculate Q1, Q3, and IQR in SQL
```{r}
run_sql("
CREATE OR REPLACE TABLE iqr_bounds AS
SELECT 
  PERCENTILE_CONT(0.25) WITHIN GROUP (ORDER BY session_numeric) AS q1,
  PERCENTILE_CONT(0.75) WITHIN GROUP (ORDER BY session_numeric) AS q3
FROM extracted_numeric
WHERE session_numeric IS NOT NULL;

SELECT * FROM iqr_bounds;
")
```

### Step 10.2: Define the upper and lower anomaly thresholds
```{r}
run_sql("
CREATE OR REPLACE TABLE anomaly_thresholds AS
SELECT 
  q1, 
  q3, 
  (q3 - q1) AS iqr,
  (q1 - 1.5 * (q3 - q1)) AS lower_bound,
  (q3 + 1.5 * (q3 - q1)) AS upper_bound
FROM iqr_bounds;

SELECT * FROM anomaly_thresholds;
")
```

---

## Step 11: Filtering Anomalies & Calculating the True Median

Now we filter out the `NULLs` and the IQR anomalies, leaving us with a statistically pure dataset to calculate our final Median.

### Step 11.1: Calculate Mean vs. Median on clean data
```{r}
run_sql("
SELECT 
  COUNT(*) AS valid_sample_size,
  ROUND(AVG(e.session_numeric), 2) AS mean_session,
  ROUND(PERCENTILE_CONT(0.5) WITHIN GROUP (ORDER BY e.session_numeric), 2) AS median_session
FROM extracted_numeric e
CROSS JOIN anomaly_thresholds t
WHERE e.session_numeric IS NOT NULL
  AND e.session_numeric BETWEEN t.lower_bound AND t.upper_bound;
")
```

**Notice the difference:** In the presence of extreme outliers and skewness, the Mean is mathematically pulled toward the tail, while the Median remains robust and represents the "typical" user.

---

## Step 12: Visualizing the Impact of Anomaly Removal in R

Let's pull the pre-cleaned and post-cleaned distributions into R to visualize why this matters.

### Step 12.1: Query raw vs cleaned distributions
```{r}
# Raw distribution (with anomalies)
raw_dist <- run_sql("
  SELECT session_numeric AS session_time, 'Raw (With Anomalies)' AS status
  FROM extracted_numeric
  WHERE session_numeric IS NOT NULL
")

# Clean distribution (anomalies removed)
clean_dist <- run_sql("
  SELECT e.session_numeric AS session_time, 'Clean (IQR Filtered)' AS status
  FROM extracted_numeric e
  CROSS JOIN anomaly_thresholds t
  WHERE e.session_numeric IS NOT NULL
    AND e.session_numeric BETWEEN t.lower_bound AND t.upper_bound
")

# Combine for plotting
plot_df <- rbind(raw_dist, clean_dist)
```

### Step 12.2: Plot the comparison
```{r}
ggplot(plot_df, aes(x = session_time, fill = status)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~status, scales = "free") +
  scale_fill_manual(values = c("Raw (With Anomalies)" = "firebrick", 
                               "Clean (IQR Filtered)" = "steelblue")) +
  labs(
    title = "The Impact of SQL Anomaly Removal on Session Time",
    subtitle = "Raw data is unreadable due to bots; IQR filtering restores the true distribution",
    x = "Session Time (Minutes)",
    y = "Density"
  ) +
  theme(legend.position = "none")
```

---

## 🏆 Conclusion: Why This Matters

| Skill | How it was demonstrated |
|-------|-------------------------|
| **String Parsing** | Used `REGEXP_REPLACE` to isolate numerics from messy strings (`"15 mins"`). |
| **Safe Casting** | Used `TRY_CAST` to prevent pipeline crashes when encountering malformed decimals (`"12.5.5"`). |
| **Statistical Outlier Detection** | Calculated Q1, Q3, and IQR entirely in SQL to define mathematical anomaly thresholds. |
| **Robust Statistics** | Demonstrated why calculating a Median is superior to a Mean when dealing with zero-inflated or bot-heavy production data. |

By structuring the SQL this way, the database does 99% of the heavy lifting. R is only used at the very end to visualize the final, pre-aggregated result. This is exactly how high-scale data pipelines operate in production.